\documentclass[11pt,a4paper]{article}

% --- Language and Font Setup ---
\usepackage[english,greek]{babel}
\usepackage{fontspec}

% Overleaf usually has 'Times New Roman' or 'FreeSerif'
\setmainfont{FreeSerif} 

% --- Math and Logic ---
\usepackage{amsmath, amssymb, amsthm}
\usepackage{listings}
\usepackage{xcolor}

% --- LOOP Language Definition ---
\lstdefinelanguage{LOOP}{
    keywords={LOOP, DO, END, IF, THEN, ELSE, :=},
    sensitive=true,
    morecomment=[l]{//},
    morestring=[b]"
}

\lstset{
    language=LOOP,
    basicstyle=\ttfamily\small,
    keywordstyle=\color{blue}\bfseries,
    commentstyle=\color{gray}\itshape,
    frame=single,
    tabsize=2,
    breaklines=true,
    numbers=left,
    numberstyle=\tiny\color{gray}
}

% --- Document Info ---
\title{Computational Models: Encoding Excersises}
\author{Vasilis Ioannou}
\date{\today}

\begin{document}

\maketitle

\section*{Encoding}

Add Here what is encoding

Cantor's Function is defined as:
\[ C(m, n) = \frac{(m+n)(m+n+1)}{2} + n \]

\subsection*{Formal Definition}
A function $C: \mathbb{N} \times \mathbb{N} \to \mathbb{N}$ must satisfy the following $m, n, k \in \mathbb{N}$:
\begin{itemize}
    \item $D_1(C(m, n)) = m$ 
    \item $D_2(C(m, n)) = n$ 
    \item $C(D_1(k), D_2(k)) = k$ 
\end{itemize}

\pagebreak
\section*{Excersise 1}

\subsection*{ Cantor Function $\text{cantor}(x, y)$}a

\begin{lstlisting}[language=LOOP]
// Input: x, y
// Result: res
sum := add(x,y)
sum_plus_one := sum + 1
prod = mult(sum,sum+plus_one)
tri = div(prod,2)
res := add(tri,y)
\end{lstlisting}

\subsection*{ Decoding Function $\text{D1}(k)$}

\begin{lstlisting}[language=LOOP]
// D1 returns the 'm' value: m = w - n
// Calculation of w = floor((sqrt(8k+1)-1)/2)
eight_k_plus_1 := mult(8,k) + 1

// Loop-based Integer SQRT
root := sqrt(eight_k_plus_1)

w := div(sub(root,1),2);
n := D2(k); // Using the result from D2
res := sub(w,n);
\end{lstlisting}

\subsection*{ Decoding Function $\text{D2}()$}

\begin{lstlisting}[language=LOOP]
// D2 returns the 'n' value: n = k - (w*(w+1)/2)
eight_k_plus_1 := mult(8,k) + 1
w := div(sub(root,1),2);

// Triangular number t = (w*(w+1))/2
w_plus_1 := w + 1;
t := div(mult(w,w_plus_1),2)
res := sub(k,t)
\end{lstlisting}

\pagebreak
\section*{Excersise 2: Closed-Form Expressions}

A \textbf{closed-form expression} is a mathematical expression that can be evaluated in a finite number of operations. It uses a standard set of well-defined functions (such as addition, multiplication, exponents, and logarithms) and avoids the use of recursion, limits, or infinite series.

\section*{The Closed-Form of the Hyperoperation Function}

The function $f(x, y, n)$ provided in the exercise is known as the \textbf{hyperoperation sequence}. While the image defines it recursively, its closed-form expression is typically represented using \textbf{Knuth's up-arrow notation} for $n \ge 3$.

The complete closed-form representation is:

\[
f(x, y, n) = 
\begin{cases} 
y + 1 & \text{if } n = 0 \\
x + y & \text{if } n = 1 \\
x \cdot y & \text{if } n = 2 \\
x \uparrow^{n-2} y & \text{if } n \ge 3 
\end{cases}
\]

\subsection*{Notation Breakdown}
\begin{itemize}
    \item \textbf{n = 3 (Exponentiation):} $x \uparrow^{3-2} y = x \uparrow y = x^y$
    \item \textbf{n = 4 (Tetration):} $x \uparrow^{4-2} y = x \uparrow\uparrow y = \underbrace{x^{x^{\cdot^{\cdot^x}}}}_{y \text{ times}}$
    \item \textbf{n = 5 (Pentation):} $x \uparrow^{5-2} y = x \uparrow\uparrow\uparrow y$
\end{itemize}

\subsection*{Excersise 3}

The set of functions computable by the LOOP language is exactly the set of Primitive Recursive Functions (PR). 
The subset of Pascal you described is simply a high-level syntax for PR.

LOOP modeling requires bounded Loops. While,Repeat and Goto modeling can have unbounded loops.
The allowed operators (+, -, *, div, mod) and logical structures (if, case) are all Primitive Recursive.
This means they can be built starting from three basic functions (See Loop Excersises)
We excluse functions and Procesures because they can allow recursion which can be unbounded.

\end{document}
